---
title: "RD_proj_KES"
format: html
editor: visual
---

## Data Processing: `plant_data` → `final_plant`

The four code blocks below reproduce the cleaning steps that transformed `Plant_Data_orig.csv` into `final_unique_plant_data.csv`.

### Step 1 — Load raw data and compare structure

```{r}
library(tidyverse)
library(stringr)
library(ggrepel)
library(scales)


plant_raw <- read_csv("Plant_Data_orig.csv")

cat("=== Plant_Data_orig.csv ===\n")
cat("Rows:", nrow(plant_raw), " | Columns:", ncol(plant_raw), "\n")
cat("Column names:\n")
print(names(plant_raw))

cat("\nUnique Filenames:", n_distinct(plant_raw$Filename),
    "vs total rows:", nrow(plant_raw), "\n")
```

### Step 2 — Normalise text to lowercase

```{r}

plant_lower <- plant_raw %>%
  mutate(across(where(is.character), str_to_lower))

cat("Before:", plant_raw$Species[1], "\n")
cat("After: ", plant_lower$Species[1], "\n")
```

### Step 3 — Drop the `Selection_Markers` column

```{r}

cat("Non-empty Selection_Markers values (before drop):",
    sum(!is.na(plant_lower$Selection_Markers) &
          plant_lower$Selection_Markers != ""), "\n")

plant_no_markers <- plant_lower %>%
  select(-Selection_Markers)

cat("Columns remaining:", ncol(plant_no_markers), "\n")
print(names(plant_no_markers))
```

### Step 4 — Deduplicate by `Filename`

```{r}
plant_clean <- plant_no_markers %>%
  group_by(Filename) %>%
  slice(1) %>%
  ungroup()

cat("Rows before deduplication:", nrow(plant_no_markers), "\n")
cat("Rows after deduplication: ", nrow(plant_clean), "\n")

# Quick comparison against the provided final file
plant_final <- read_csv("final_unique_plant_data.csv")
cat("\nRow count matches final file:",
    nrow(plant_clean) == nrow(plant_final), "\n")
cat("Column count matches final file (excluding unnamed col):",
    ncol(plant_clean) == ncol(plant_final) - 1L, "\n")
```

------------------------------------------------------------------------

```{r}
library(tidyverse)
library(stringr)
library(ggrepel)
library(scales)

plant <- read_csv("final_unique_plant_data.csv")
glimpse(plant)
```

Objective 3: Map the relationship between specific gene families and high-value agricultural traits like drought tolerance and disease resistance.

Creating trait labels and infer gene family

```{r}
obj3 <- plant %>%
  mutate(
    trait_group = case_when(
      str_detect(str_to_lower(Application_Category), "drought tolerance") |
        str_detect(str_to_lower(Phenotype), "drought") ~ "Drought tolerance",

      str_detect(str_to_lower(Application_Category), "disease resistance") |
        str_detect(str_to_lower(Phenotype), "disease|blight|fungal|pathogen|viral|virus|resistance") ~ "Disease resistance",

      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(trait_group)) %>%
  mutate(
    Gene_Target = str_to_lower(Gene_Target),

    # keep only first listed target if multiple are present
    Gene_Target = str_split(Gene_Target, ",|;|/|\\|") %>% map_chr(1),
    Gene_Target = str_trim(Gene_Target),

    # remove common species prefixes
    Gene_Target_clean = str_remove(Gene_Target,
      "^(at|os|sl|zm|ta|hv|st|gh|gm|mt|rc|cs|le|pe|ib|pg|md|pa|vv|bn|bna|loc|loc_)[-_]?"
    ),

    gene_family = case_when(
      str_detect(Gene_Target_clean, "^myb") ~ "MYB",
      str_detect(Gene_Target_clean, "^nac") ~ "NAC",
      str_detect(Gene_Target_clean, "^wrky") ~ "WRKY",
      str_detect(Gene_Target_clean, "^dreb") ~ "DREB",
      str_detect(Gene_Target_clean, "^erf|^ap2") ~ "AP2/ERF",
      str_detect(Gene_Target_clean, "^bzip") ~ "bZIP",
      str_detect(Gene_Target_clean, "^mlo") ~ "MLO",
      str_detect(Gene_Target_clean, "^sweet") ~ "SWEET",
      str_detect(Gene_Target_clean, "r genes|r gene|r-gene") ~ "R genes",
      str_detect(Gene_Target_clean, "^pr") ~ "PR",
      TRUE ~ str_extract(str_to_upper(Gene_Target_clean), "^[A-Z]+")
    )
  ) %>%
  filter(!is.na(gene_family), gene_family != "")
```

Summary of top gene families

```{r}
obj3_summary <- obj3 %>%
  count(gene_family, trait_group, sort = TRUE) %>%
  group_by(gene_family) %>%
  mutate(total_family_n = sum(n)) %>%
  ungroup() %>%
  filter(total_family_n >= 3)   # adjust threshold
obj3_summary
```

Heatmap of gene family vs traits

```{r}
heatmap_plot <- obj3_summary %>%
  ggplot(aes(
    x = trait_group,
    y = fct_reorder(gene_family, total_family_n),
    fill = n
  )) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "grey90", high = "steelblue") +
  labs(
    title = "Gene family associations with drought tolerance and disease resistance",
    x = "Trait category",
    y = "Gene family",
    fill = "Count of records"
  ) +
  theme_minimal(base_size = 12)

heatmap_plot
```

Bubble plot

```{r}
bubble_plot <- obj3_summary %>%
  ggplot(aes(
    x = trait_group,
    y = fct_reorder(gene_family, total_family_n),
    size = n
  )) +
  geom_point(alpha = 0.75) +
  scale_size_continuous(range = c(3, 12)) +
  labs(
    title = "Strength of gene family–trait relationships",
    x = "Trait category",
    y = "Gene family",
    size = "Count of records"
  ) +
  theme_minimal(base_size = 12)

bubble_plot
```

Objective 4 (Quantitative Analysis):

Extract and analyze numerical data from Quant_Data text to evaluate whether CRISPR-edited plants demonstrate greater percentage improvements in yield compared to traditional transgenic plants.

define the technology groups

```{r}
obj4_simple <- plant %>%
  mutate(
    tech_group = case_when(
      str_detect(str_to_lower(Mod_type), "crispr|cas9|cas12|base editing|prime editing") ~ "CRISPR",
      str_detect(str_to_lower(Mod_type), "transgenic|overexpression|agrobacterium|rnai|particle bombardment|t-dna|transformation") ~ "Transgenic",
      TRUE ~ NA_character_
    ),
    yield_related =
      str_detect(str_to_lower(Application_Category), "yield") |
      str_detect(str_to_lower(Phenotype), "yield") |
      str_detect(str_to_lower(Quant_Data), "yield"),

    percent_match = str_extract(Quant_Data, "\\d+\\.?\\d*\\s*%"),
    yield_pct = as.numeric(str_remove(percent_match, "%"))
  ) %>%
  filter(!is.na(tech_group), yield_related, !is.na(yield_pct))
```

summary

```{r}
obj4_summary <- obj4_simple %>%
  group_by(tech_group) %>%
  summarise(
    n_values = n(),
    mean_pct = mean(yield_pct, na.rm = TRUE),
    median_pct = median(yield_pct, na.rm = TRUE),
    sd_pct = sd(yield_pct, na.rm = TRUE),
    min_pct = min(yield_pct, na.rm = TRUE),
    max_pct = max(yield_pct, na.rm = TRUE)
  )

obj4_summary
```

Boxplot

```{r}
ggplot(obj4_simple, aes(x = tech_group, y = yield_pct)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(
    title = "Yield improvement in CRISPR vs Transgenic plants",
    x = "Modification type",
    y = "Yield change (%)"
  ) +
  theme_minimal()
```

bar chart

```{r}
ggplot(obj4_summary, aes(x = tech_group, y = mean_pct)) +
  geom_col() +
  labs(
    title = "Average yield improvement by modification type",
    x = "Modification type",
    y = "Mean yield change (%)"
  ) +
  theme_minimal()
```

stat test

```{r}

wilcox.test(yield_pct ~ tech_group, data = obj4_simple)
```
