---
title: "RD_proj_KES"
format: html
editor: visual
---

## Data Cleaning: `Plant_Data_orig.csv` → `final_unique_plant_data.csv`

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(ggrepel)
library(scales)

plant_raw <- read_csv("Plant_Data_orig.csv")
```

### Step 1 — Remove incomplete records (empty Gene rows)

```{r}
plant_raw %>%
  filter(is.na(Gene) | Gene == "") %>%
  select(Filename, Species, Gene, Mod_Type) %>%
  print()

plant_step1 <- plant_raw %>%
  filter(!is.na(Gene), Gene != "")

cat("Rows removed (no Gene):", nrow(plant_raw) - nrow(plant_step1), "\n")
cat("Rows remaining:        ", nrow(plant_step1), "\n")
```

### Step 2 — One row per paper: keep highest-relevance gene

The raw file has one row per gene per paper. the `Relevance_Score` column tells us which gene row best represents the paper's main finding

```{r}

plant_step1 %>%
  filter(Filename == "9087934.pdf") %>%
  select(Filename, Relevance_Score, Gene, Mod_Type, Phenotype) %>%
  print()

plant_step2 <- plant_step1 %>%
  group_by(Filename) %>%
  slice_max(Relevance_Score, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Rows before deduplication:", nrow(plant_step1), "\n")
cat("Rows after deduplication: ", nrow(plant_step2), "\n")
cat("Unique papers:            ", n_distinct(plant_step2$Filename), "\n")
```

### Step 3 — Drop `Relevance_Score` and `Terminator`

```{r}

cat("Non-empty Terminator cells:",
    sum(!is.na(plant_step2$Terminator) & plant_step2$Terminator != ""), "\n")

plant_step3 <- plant_step2 %>%
  select(-Relevance_Score, -Terminator)

cat("Columns after drop:", ncol(plant_step3), "\n")
print(names(plant_step3))
```

### Step 4 — Rename `Gene` → `Gene_Target` and `Mod_Type` → `Mod_type`

```{r}
plant_step4 <- plant_step3 %>%
  rename(
    Gene_Target = Gene,
    Mod_type    = Mod_Type
  )

cat("Renamed columns:\n")
print(names(plant_step4))
```

### Step 5 — Derive `Mod_Method` from `Mod_type`

```{r}
plant_step5 <- plant_step4 %>%
  mutate(
    Mod_Method = case_when(
      str_detect(str_to_lower(Mod_type),
        "agrobacterium|agro.infiltration")                         ~ "Agrobacterium",
      str_detect(str_to_lower(Mod_type),
        "crispr|cas9|cas12|base edit|prime edit")                  ~ "CRISPR/Cas9",
      str_detect(str_to_lower(Mod_type),
        "biolistic|bombardment|particle gun")                      ~ "Biolistics",
      str_detect(str_to_lower(Mod_type),
        "rnai|rna interference|sirna|shrna|antisense|gene silenc|mirna|mir159") ~ "RNAi",
      str_detect(str_to_lower(Mod_type),
        "vigs|virus.induced gene silenc")                          ~ "VIGS",
      TRUE ~ "Unknown"
    )
  )

cat("Mod_Method distribution:\n")
print(table(plant_step5$Mod_Method))
```

### Step 6 — Derive `Mod_Direction` from `Mod_type`

```{r}
plant_step6 <- plant_step5 %>%
  mutate(
    Mod_Direction = case_when(
      str_detect(str_to_lower(Mod_type),
        "overexpression|overexpress|upregulat|up-regulat|transcriptional activation|promoter replacement") ~ "Overexpression",
      str_detect(str_to_lower(Mod_type),
        "knockout|knockdown|knock.out|knock.down|deletion|gene silenc|silencing|rnai|downregulat|down-regulat|suppression|loss.of.function|transcriptional repression|antisense") ~ "Knockdown/Out",
      str_detect(str_to_lower(Mod_type),
        "mutation|mutant|mutagenesis|snp|snv|indel|missense|insertional mutant|inversion|natural mutation") ~ "Mutation",
      TRUE ~ "Modification"
    )
  )

cat("Mod_Direction distribution:\n")
print(table(plant_step6$Mod_Direction))
```

### Step 7 — Standardise text: lowercase + collapse whitespace

```{r}

plant_step6 %>%
  count(Application_Category, sort = TRUE) %>%
  head(6) %>%
  print()

plant_step7 <- plant_step6 %>%
  mutate(across(where(is.character), ~ str_squish(str_to_lower(.x))))


plant_step7 %>%
  count(Application_Category, sort = TRUE) %>%
  head(6) %>%
  print()
```

### Step 8 — Extract publication year from DOI

DOIs frequently encode the publication year as a four-digit number in the suffix (e.g. `10.1186/s12870-**2022**-03584-y`). We pull the first match of `(19|20)\d{2}` and store it as an integer column `Pub_Year`.

```{r}
current_year <- as.integer(format(Sys.Date(), "%Y"))

plant_step8 <- plant_step7 %>%
  mutate(
    # (?<![0-9]) and (?![0-9]) ensure the 4 digits are NOT part of a
    # longer number (e.g. avoids matching "2099" inside "112099")
    Pub_Year = str_extract(DOI, "(?<![0-9])(19|20)\\d{2}(?![0-9])") %>%
                 as.integer(),
    # sanity-check: keep only plausible publication years
    Pub_Year = if_else(Pub_Year >= 1990 & Pub_Year <= current_year,
                       Pub_Year, NA_integer_)
  )

cat("Year extraction summary:\n")
cat("  Papers with year found :", sum(!is.na(plant_step8$Pub_Year)), "\n")
cat("  Papers without year    :", sum( is.na(plant_step8$Pub_Year)), "\n")
cat("  Year range             :",
    min(plant_step8$Pub_Year, na.rm = TRUE), "–",
    max(plant_step8$Pub_Year, na.rm = TRUE), "\n\n")

plant_step8 %>%
  count(Pub_Year, sort = FALSE) %>%
  print(n = Inf)
```

------------------------------------------------------------------------

```{r}
# Use the fully cleaned pipeline output directly
plant <- plant_step8
glimpse(plant)
```

Objective 1: Trends in gene-editing technologies over time

```{r}
# Pub_Year already extracted in Step 8 — no inline re-extraction needed
obj1 <- plant %>%
  filter(!is.na(Pub_Year)) %>%
  mutate(
    tech_type = case_when(
      str_detect(Mod_type, "crispr|cas9|cas12|base editing|prime editing") ~ "CRISPR",
      str_detect(Mod_type, "rnai")                                          ~ "RNAi",
      str_detect(Mod_type,
        "transgenic|overexpression|agrobacterium|t-dna|transformation|particle bombardment") ~ "Transgenic",
      TRUE ~ "Other"
    )
  )

# yearly counts
obj1_summary <- obj1 %>%
  count(Pub_Year, tech_type)

obj1_summary
```

```{r}
ggplot(obj1_summary, aes(x = Pub_Year, y = n, color = tech_type)) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  labs(
    title = "Evolution of gene-editing technologies in plant biology",
    x = "Publication Year",
    y = "Number of studies",
    color = "Technology"
  ) +
  theme_minimal()
```

When did CRISPR dominate?

```{r}
crispr_dominance <- obj1_summary %>%
  group_by(Pub_Year) %>%
  mutate(total = sum(n)) %>%
  filter(tech_type == "CRISPR") %>%
  mutate(prop = n / total) %>%
  arrange(Pub_Year)

crispr_dominance
```

Objective 2: Plant family focus and under-studied crops

```{r}
obj2 <- plant %>%
  mutate(
    species_lower = str_to_lower(Species),

    plant_family = case_when(
      str_detect(species_lower, "rice|wheat|maize|corn|barley|sorghum") ~ "Poaceae",
      str_detect(species_lower, "tomato|potato|pepper|eggplant") ~ "Solanaceae",
      str_detect(species_lower, "soybean|bean|pea|lentil") ~ "Fabaceae",
      str_detect(species_lower, "arabidopsis") ~ "Brassicaceae",
      str_detect(species_lower, "cotton") ~ "Malvaceae",
      TRUE ~ "Other"
    )
  )
```

Summary counts

```{r}
obj2_summary <- obj2 %>%
  count(plant_family, sort = TRUE)

obj2_summary
```

```{r}
ggplot(obj2_summary, aes(x = fct_reorder(plant_family, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Plant families studied in functional genomics",
    x = "Plant family",
    y = "Number of studies"
  ) +
  theme_minimal()
```

Identify under-studied species

```{r}
species_counts <- obj2 %>%
  count(Species, sort = TRUE)

understudied <- species_counts %>%
  filter(n <= 2)   # threshold can be adjusted

understudied
```

under-studied vs major crops

```{r}
obj2 %>%
  mutate(category = if_else(Species %in% understudied$Species, "Under-studied", "Well-studied")) %>%
  count(category) %>%
  ggplot(aes(x = category, y = n, fill = category)) +
  geom_col() +
  labs(
    title = "Distribution of study focus across plant species",
    x = "",
    y = "Count"
  ) +
  theme_minimal()
```

Objective 3: Map the relationship between specific gene families and high-value agricultural traits like drought tolerance and disease resistance.

Creating trait labels and infer gene family

```{r}
obj3 <- plant %>%
  mutate(
    trait_group = case_when(
      str_detect(str_to_lower(Application_Category), "drought tolerance") |
        str_detect(str_to_lower(Phenotype), "drought") ~ "Drought tolerance",

      str_detect(str_to_lower(Application_Category), "disease resistance") |
        str_detect(str_to_lower(Phenotype), "disease|blight|fungal|pathogen|viral|virus|resistance") ~ "Disease resistance",

      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(trait_group)) %>%
  mutate(
    Gene_Target = str_to_lower(Gene_Target),

    # keep only first listed target if multiple are present
    Gene_Target = str_split(Gene_Target, ",|;|/|\\|") %>% map_chr(1),
    Gene_Target = str_trim(Gene_Target),

    # remove common species prefixes
    Gene_Target_clean = str_remove(Gene_Target,
      "^(at|os|sl|zm|ta|hv|st|gh|gm|mt|rc|cs|le|pe|ib|pg|md|pa|vv|bn|bna|loc|loc_)[-_]?"
    ),

    gene_family = case_when(
      str_detect(Gene_Target_clean, "^myb") ~ "MYB",
      str_detect(Gene_Target_clean, "^nac") ~ "NAC",
      str_detect(Gene_Target_clean, "^wrky") ~ "WRKY",
      str_detect(Gene_Target_clean, "^dreb") ~ "DREB",
      str_detect(Gene_Target_clean, "^erf|^ap2") ~ "AP2/ERF",
      str_detect(Gene_Target_clean, "^bzip") ~ "bZIP",
      str_detect(Gene_Target_clean, "^mlo") ~ "MLO",
      str_detect(Gene_Target_clean, "^sweet") ~ "SWEET",
      str_detect(Gene_Target_clean, "r genes|r gene|r-gene") ~ "R genes",
      str_detect(Gene_Target_clean, "^pr") ~ "PR",
      TRUE ~ str_extract(str_to_upper(Gene_Target_clean), "^[A-Z]+")
    )
  ) %>%
  filter(!is.na(gene_family), gene_family != "")
```

Summary of top gene families

```{r}
obj3_summary <- obj3 %>%
  count(gene_family, trait_group, sort = TRUE) %>%
  group_by(gene_family) %>%
  mutate(total_family_n = sum(n)) %>%
  ungroup() %>%
  filter(total_family_n >= 3)   # adjust threshold
obj3_summary
```

Heatmap of gene family vs traits

```{r}
heatmap_plot <- obj3_summary %>%
  ggplot(aes(
    x = trait_group,
    y = fct_reorder(gene_family, total_family_n),
    fill = n
  )) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "grey90", high = "steelblue") +
  labs(
    title = "Gene family associations with drought tolerance and disease resistance",
    x = "Trait category",
    y = "Gene family",
    fill = "Count of records"
  ) +
  theme_minimal(base_size = 12)

heatmap_plot
```

Bubble plot

```{r}
bubble_plot <- obj3_summary %>%
  ggplot(aes(
    x = trait_group,
    y = fct_reorder(gene_family, total_family_n),
    size = n
  )) +
  geom_point(alpha = 0.75) +
  scale_size_continuous(range = c(3, 12)) +
  labs(
    title = "Strength of gene family–trait relationships",
    x = "Trait category",
    y = "Gene family",
    size = "Count of records"
  ) +
  theme_minimal(base_size = 12)

bubble_plot
```

Objective 4 (Quantitative Analysis):

Extract and analyze numerical data from Quant_Data text to evaluate whether CRISPR-edited plants demonstrate greater percentage improvements in yield compared to traditional transgenic plants.

define the technology groups

```{r}
obj4_simple <- plant %>%
  mutate(
    tech_group = case_when(
      str_detect(str_to_lower(Mod_type), "crispr|cas9|cas12|base editing|prime editing") ~ "CRISPR",
      str_detect(str_to_lower(Mod_type), "transgenic|overexpression|agrobacterium|rnai|particle bombardment|t-dna|transformation") ~ "Transgenic",
      TRUE ~ NA_character_
    ),
    yield_related =
      str_detect(str_to_lower(Application_Category), "yield") |
      str_detect(str_to_lower(Phenotype), "yield") |
      str_detect(str_to_lower(Quant_Data), "yield"),

    percent_match = str_extract(Quant_Data, "\\d+\\.?\\d*\\s*%"),
    yield_pct = as.numeric(str_remove(percent_match, "%"))
  ) %>%
  filter(!is.na(tech_group), yield_related, !is.na(yield_pct))
```

summary

```{r}
obj4_summary <- obj4_simple %>%
  group_by(tech_group) %>%
  summarise(
    n_values = n(),
    mean_pct = mean(yield_pct, na.rm = TRUE),
    median_pct = median(yield_pct, na.rm = TRUE),
    sd_pct = sd(yield_pct, na.rm = TRUE),
    min_pct = min(yield_pct, na.rm = TRUE),
    max_pct = max(yield_pct, na.rm = TRUE)
  )

obj4_summary
```

Boxplot

```{r}
ggplot(obj4_simple, aes(x = tech_group, y = yield_pct)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(
    title = "Yield improvement in CRISPR vs Transgenic plants",
    x = "Modification type",
    y = "Yield change (%)"
  ) +
  theme_minimal()
```

bar chart

```{r}
ggplot(obj4_summary, aes(x = tech_group, y = mean_pct)) +
  geom_col() +
  labs(
    title = "Average yield improvement by modification type",
    x = "Modification type",
    y = "Mean yield change (%)"
  ) +
  theme_minimal()
```

stat test

```{r}
wilcox.test(yield_pct ~ tech_group, data = obj4_simple)
```
